Expert-In-The-Loop Causal Discovery: \\ Iterative Model Refinement Using Expert Knowledge

### 2025-02-10 paper/main.tex
Empirical section

\subsection{Effect Sizes for CI Tests}

@@

In this section, we give some of the practical aspects of the
\textsc{ExpertInLoop} algorithm, and then compare the algorithm with automated
causal discovery algorithms.

In all experiments, we use the \textsc{RankedExpand} method together
with a version of the strong ancestral oracle that can make mistakes.

\todo{Want to know: what if my experts are not perfect?}

The \textsc{ExpertInLoop} algorithm requires two oracles, one that can answer
d-separation queries and an ancestral oracle that can give the causal direction
between variable pairs. For the experiments, we simulate the d-separation
oracle using a CI test with a p-value threshold of $ 0.05 $. To simulate the
edge orientation oracle, we simulate a domain expert that can give the correct
causal relationship with some accuracy $ \alpha $.

@@

An important point to note here is that the effective accuracy of $
\mathrm{Expert} $ is higher than $ \alpha $ as even when giving a random
answer, there is a $ 1 $ in $ 3 $ chance that the answer is correct by chance.
Hence, for an $ \mathrm{Expert} $ with accuracy $ \alpha $, the effective
accuracy becomes:

@@

We analyzed the performance of \textsc{ExpertInLoop} algorithm by comparing it
with three other automated algorithms: PC, Hill-Climb Search, and Greedy
Equivalence Search (GES). For the analysis, we simulated data from known DAGs
and use the algorithms to recover the original DAG. We start by generating a
random DAG on $10$ nodes and use linear models with random effects to simulate
$ 500 $ samples from it. We simulate each variable $ V_i $ in the DAG as:

\begin{equation}
	\begin{split}
		\beta &= \mathrm{Uniform}([-0.6, 0.6]) \\
		V_i &= \bm{\beta} \mathrm{Pa}_G(V_i) + \mathcal{N}(0, 1)
	\end{split}
\end{equation}

@@

To compare the learned DAG to the original DAG, we use two metrics: Structural
Hamming Distance (SHD) and Structural Intervention Distance (SID)
\citep{Peters2015}. Unlike \textrm{ExpertInLoop}, the automated algorithms can
only recover the CPDAG, we consider all possible orientations of the CPDAG to
compute the range of SHD and SID values to show the best and worst case
results. The results of this analysis are shown in Figure~\ref{fig:shd_sid}.
For SHD, the performance of \textsc{ExpertInLoop} is comparable to the
automated algorithms for $ \alpha = 0.3 (\alpha_{\textrm{eff}} = 0.53 $, which
is equivalent to the expert getting the orientation correct $ 1 $ out of $ 2 $
case. For SID, the performance is comparable to the automated algorithms for $
\alpha = 0.5 (\alpha_{\textrm{eff}} = 0.67) $, which is equivalent to the
expert getting the orientation correct in $ 2 $ out of $ 3 $ cases. For denser
DAGs, the \textsc{ExpertInLoop} is able to perform better for lower $ \alpha $
values.

>>>>

\subsection{Effect Sizes for CI Tests}
\label{sec:ci_tests}

@@

In this section, we compare our \textsc{ExpertInLoop} algorithm with automated
causal discovery algorithms. Our goal with the empirical analysis is to
understand the behaviour of our method when the experts are not perfect. To
simulate this imperfect expert, we use a version of the strong ancestral oracle
that has an accuracy of $ \alpha $:

@@

An important point to note here is that the effective accuracy of $
\mathrm{Expert} $ is higher than $ \alpha $ as even when giving a random
answer, there is a $ 1 $ in $ 3 $ chance that the answer is correct. Therefore,
an $ \mathrm{Expert} $ with accuracy $ \alpha $, has an effective accuracy:

@@

, we use the
\textsc{RankedExpand} method together with a version of the strong ancestral
oracle that can make mistakes.

We analyzed the performance of \textsc{ExpertInLoop} algorithm by comparing it
with three other automated algorithms: PC, Hill-Climb Search, and Greedy
Equivalence Search (GES) on linear Gaussian data. For \textsc{ExpertInLoop}, we
use the \textsc{RankedExpand} method with Pearson's correlation coefficient as
the effect size measure (Section~\ref{ref:ci_tests}). For the analysis, we
simulated data from known DAGs and use the algorithms to recover the original
DAG. We start by generating a random DAG on $10$ nodes and use linear models
with random effects to simulate $ 500 $ samples from it. We simulate each
variable $ V_i $ in the DAG as:

\begin{equation}
	\begin{split}
		\bm{\beta} &= \mathrm{Uniform}([-0.6, 0.6]) \\
		V_i &= \bm{\beta} \mathrm{Pa}_G(V_i) + \mathcal{N}(0, 1)
	\end{split}
\end{equation}

@@

To compare the learned DAG to the original DAG, we use two metrics: Structural
Hamming Distance (SHD) and Structural Intervention Distance (SID)
\citep{Peters2015}. Unlike \textrm{ExpertInLoop}, the automated algorithms can
only recover the MEC, therefore, we considered all possible orientations of the
MEC to compute a range of SHD and SID values representing the best and worst
case scenarios. The results of this analysis are shown in
Figure~\ref{fig:shd_sid}. For SHD, the performance of \textsc{ExpertInLoop} is
comparable to the automated algorithms for $ \alpha = 0.3
(\alpha_{\textrm{eff}} = 0.53 $, implying that if an expert is able to get the
edge orientation correctly in more than $ 1 $ out of $ 2 $ cases, they can
outperform the automated algorithms. Similarly for SID, an expert can outperform 
the automated algorithms if they are able to get the orientation correct in $ 2 $ 
out of $ 3 $ cases. Another thing to note is that for denser DAGs, the 
\textsc{ExpertInLoop} is able to perform better for lower $ \alpha $ values.
